A Dearth of Young Stars and Planets in the Kepler Field

### 2024-04-26 paper/ms.tex
update ms

Exoplanet science is in an age of abundance, fueled by the discovery
of thousands of worlds orbiting close to their host stars
\citep{Borucki10,2015JATIS...1a4003R}.
Most of these exoplanets have ages comparable to that of the solar
system.
This fact leaves many gaps in our knowledge of what exact dynamical
and physical processes produce these objects.
Finding and characterizing young ($\lesssim$1\,Gyr) exoplanets offers
one possible workaround.
The reason is that processes such as thermal cooling, atmospheric
loss, and dynamical instability occur over $\approx$10\,Myr to
$\approx$1\,Gyr timescales in a way that can leave observable imprints in
the properties of young planets (CITE theory reviews).
Individual exemplars include V1298~Tau \citep{David_2019}, an
adolescent resonant chain of four transiting puffy planets,
and HIP~67522 \citep{Rizzuto_2020}, a Jupiter-sized planet with a
Neptunian mass (CITE).
Population-level examples include tentative differences in the size
distribution of young vs.~old planets
(CITEVach,CITEChristianse,CiteDavid).

@@

\section{Selecting the Stellar and Planetary Samples}
\label{sec:selection}

\begin{figure*}[!t]
	\begin{center}
		\subfloat{
			\includegraphics[width=0.47\textwidth]{st_params_M_G_vs_dr3_bp_rp.pdf}
		}
%	
%		\subfloat{
%			\includegraphics[width=0.4\textwidth]{st_params_M_G_vs_adopted_Teff.pdf}
%		}
%		\vspace{-0.5cm}
		\subfloat{
		\includegraphics[width=0.47\textwidth]{st_params_adopted_logg_vs_adopted_Teff.pdf}
		}
%		\subfloat{
%			\includegraphics[width=0.5\textwidth]{st_params_dr3_phot_g_mean_mag_vs_adopted_Teff.pdf}
%		}
	\end{center}
	\vspace{-0.5cm}
	\caption{
    {\bf The stars.}  Our analysis focuses on stars observed
    by Kepler with previously reported rotation periods (dark gray
    points).  About half of these rotators are suitable for
    gyrochronology (black points), based on factors including their
    temperatures, surface gravities, (non-)binarity, and proximity to
    the main sequence.  The full list of criteria is given in
    Section~\ref{subsec:flags}.  A few percent of the remainder are
    either ``confirmed'' or ``candidate'' Kepler Objects of Interest
    that meet an additional set of planetary quality criteria (blue points;
    Section~\ref{subsec:plflags}).  
%	We independently derive
%    lithium-based ages for a largely-overlapping setting of KOIs based
%    on high-resolution spectroscopy.
    Surface gravities and effective
    temperatures are primarily from
    \citet{Berger_2020a_catalog}.
	}
	\label{fig:stellarprops}
\end{figure*}

@@

The three main assumptions required for a rotation-based age dating
method like \texttt{gyro-interp} to be valid are as follows.  {\it 1)}
The target star must be near the main sequence, {\it 2)} its spindown
must not be influenced by binary companions, and {\it 3)} the rotation
period distribution for field stars of a given mass and age must be
identical to that of equally aged open clusters (metallicity
differences, for instance, are ignored).  Regarding the former two
assumptions: pre-main-sequence contraction and post-main-sequence
expansion influence surface stellar rotation periods in ways that we
do not model.  Stars in e.g.~ close binaries might be subject to tidal
effects that we do not treat.

Using our adopted effective temperatures and rotation periods, we ran
\texttt{gyro-interp} over an age grid of 0 to 4\,Gyr, linearly spaced
over 500 grid points.  \texttt{gyro-interp} itself is calibrated
out to 4\,Gyr, the adopted age of the M67 cluster
\citep[see][]{2022ApJ...938..118D,Gruner_2023}.  To estimate the stellar
spin-down rate as a function of time and stellar temperature, the
framework interpolates between reference open clusters using smooth
cubic hermite polynomials.  We calculate the probability of the
rotation-based age $t_{\rm gyro}$, given the observed rotation
periods, effective temperatures, and their uncertainties by
integrating Equation~1 of \citet{Bouma_2023} at the default grid
resolution.  This procedure marginalizes over the intrinsic age and
temperature dependent scatter that is observed in the cluster
sequences to yield statistical age uncertainties.

@@

@@

{\it Absolute luminosity (bit 2)}---We calculated the absolute Gaia
DR3 $G$-band luminosity, ignoring reddening, using the reported
apparent $G$-band magnitude and parallaxes.  We flag stars with
$M_{\rm G} < 3.9$, corresponding to main-sequence spectral types
earlier than $\approx$F8V \citep{Pecaut_2013}.

@@

{\it RUWE (bit 7)}---We flagged stars with a Gaia DR3 renormalized
unit weight error
(RUWE) %\footnote{\url{https://gea.esac.esa.int/archive/documentation/GDR3/Gaia_archive/chap_datamodel/sec_dm_main_source_catalogue/ssec_dm_gaia_source.html}})
exceeding 1.4 as possible binaries.  Such astrometric outliers can be
either bona fide astrometric binaries, or more often are marginally
resolved point sources for which the single-source PSF model used in
the Gaia pipeline induces spurious astrometric scatter.

@@

{\it Near the main-sequence (bit 9)}---Many stars observed by Kepler
are far from the main sequence: this can be seen in color--absolute
magnitude diagrams, and in diagrams of e.g.~surface gravity as a
function of temperature (Figure~\ref{fig:stellarprops}).  This is a
challenge for rotation-based age-dating because {\it i)} point sources
far from the main-sequence include unresolved photometric binaries,
and {\it ii}) effects other than magnetic braking influence the
rotation periods of bona fide single stars far from the main-sequence.
The added complications of reddening and metallicity make defining a
simple cut in any two dimensions challenging.  We explored various
options, and found that the $+1$$\sigma$ relative isochronal
uncertainty reported by \citet{Berger_2020a_catalog}, $+\sigma_{t,{\rm
B20,iso}}/t_{\rm B20,iso}$, reasonably separated FGK stars ``near'' and
``far'' from the main-sequence.  We therefore selected stars with
$t_{\rm B20,iso}/(+\sigma_{t,{\rm B20,iso}})\approx 3$, and fitted an
an $N^{\rm th}$-order polynomial in the $\log g$-$T_{\rm eff}$ plane
to these stars between 4100 and 5800\,K.  We let the order of the fit
vary from $N$=1 to 10, and minimizing the Bayesian Information Criterion
yielded $N$=6.  The resulting locus is shown in
Figure~\ref{fig:stellarprops}.  While the details of this procedure
are entirely arbitrary, the overall purpose is to exclude stars for
which there is photometric evidence that gyrochronology may not be
reliable.  The rotation period distribution of sources excluded in
this manner includes a larger fraction of ``rapid rotators'', likely
due to a combination of astrophysical and systematic effects.

@@

An important decision for our subsequent analysis is how to combine
the above flags in order to decide whether or not a star is suited for
gyrochronology.  By default, we assume that a star is suitable for
gyrochronology if none of bits zero through nine (inclusive) are
raised.

@@

{\it Candidate S/N (bit 1)}---the false positive rate increases
greatly toward Kepler's noise floor for planet detection (CITE).  We
required a S/N in excess of the usual floor, through a cut on the
``multiple event statistic'' (MES): we required ${\rm MES}$$>$10.

@@

In the planet sample, the cumulative KOI table included
\nnonfpcumkois\ confirmed and candidate planets, orbiting
\nnonfpcumkoihosts\ stars.  Assume that we apply all the same stellar
cuts, save for that on RUWE.  Then, require these confirmed and
candidate planets to have MES$>$10.  This combination of selection
criteria yield \nplwgyroagewithgrazingandhighruwe\ planets with
rotation-based ages orbiting
\nplhoststarwgyroagewithgrazingandhighruwe\ stars
(\nplhoststarwgyroagejusthighruwe\ of these stars with ``high'' RUWE).
If we additionally require the planets to be non-grazing ($b<0.8$)
based on the Markov Chain Monte Carlo (MCMC) fits from
\citet{Thompson_2018}, we are left with \nplwgyroagenograzing\ and
\nplhoststarwgyroagenograzing\ planets and stars, respectively.

>>>>

Exoplanet science is in an age of abundance, fueled by the discovery
of thousands of worlds orbiting close to their host stars
\citep{Borucki10,2015JATIS...1a4003R}.
Most of these exoplanets have ages comparable to that of the solar
system.
This fact leaves many gaps in our knowledge of the exact dynamical
and physical processes that produce these objects.
Finding and characterizing young ($\lesssim$1\,Gyr) exoplanets offers
one possible workaround.
The reason is that processes such as thermal cooling, atmospheric
loss, and dynamical instability occur over $\approx$10\,Myr to
$\approx$1\,Gyr timescales, and so they can leave observable imprints in
the properties of young planets (CITE theory reviews).
Individual exemplars include V1298~Tau \citep{David_2019}, an
adolescent resonant chain of four transiting puffy planets,
and HIP~67522 \citep{Rizzuto_2020}, a Jupiter-sized planet with a
Neptunian mass (CITE).
% And perhaps also TOI837, now with a uber massive core!
Population-level examples include tentative differences in the size
distribution of young vs.~old planets
(CITEVach,CITEChristianse,CiteDavid).

@@

\section{Selecting the Stars and Planets}
\label{sec:selection}

\begin{figure*}[!t]
	\begin{center}
		\subfloat{
			\includegraphics[width=0.49\textwidth]{st_params_M_G_vs_dr3_bp_rp.pdf}
		}
%	
%		\subfloat{
%			\includegraphics[width=0.4\textwidth]{st_params_M_G_vs_adopted_Teff.pdf}
%		}
%		\vspace{-0.5cm}
		\subfloat{
		\includegraphics[width=0.49\textwidth]{st_params_adopted_logg_vs_adopted_Teff.pdf}
		}
%		\subfloat{
%			\includegraphics[width=0.5\textwidth]{st_params_dr3_phot_g_mean_mag_vs_adopted_Teff.pdf}
%		}
	\end{center}
	\vspace{-0.5cm}
	\caption{
    {\bf The stars.}  Our analysis focuses on stars observed
    by Kepler with previously reported rotation periods (dark gray
    points).  About half of these rotators are suitable for
    gyrochronology (black points), based on factors including their
    temperatures, surface gravities, (non-)binarity, and proximity to
    the main sequence.  The full list of criteria is given in
    Section~\ref{subsec:flags}.  A few percent of the remainder are
    either ``confirmed'' or ``candidate'' Kepler Objects of Interest
    that meet an additional set of planetary quality criteria (blue points;
    Section~\ref{subsec:plflags}).  
%	We independently derive
%    lithium-based ages for a largely-overlapping setting of KOIs based
%    on high-resolution spectroscopy.
    Surface gravities and effective
    temperatures are primarily from
    \citet{Berger_2020a_catalog}.
	}
	\label{fig:stellarprops}
\end{figure*}

@@

Using our adopted effective temperatures and rotation periods, we ran
\texttt{gyro-interp} over an age grid of 0 to 4\,Gyr, linearly spaced
over 500 grid points.  \texttt{gyro-interp} itself is calibrated out
to 4\,Gyr, the age of M67
\citep[see][]{2022ApJ...938..118D,Gruner_2023}.  To estimate the
stellar spin-down rate as a function of time and stellar temperature,
the framework interpolates between reference open clusters using
smooth cubic hermite polynomials.  We calculate the probability of the
rotation-based age $t_{\rm gyro}$, given the observed rotation
periods, effective temperatures, and their uncertainties by
integrating Equation~1 of \citet{Bouma_2023} at the default grid
resolution.  This procedure marginalizes over the intrinsic age and
temperature dependent scatter that is observed in the cluster
sequences to yield statistical age uncertainties.

@@

As a general philosophy, three assumptions must hold for a
rotation-based age dating method like \texttt{gyro-interp} to be
valid.  {\it 1)} The evolutionary state of the target star must be
well-specified by its temperature and its age, {\it 2)} the star's
spindown must not be influenced by binary companions, and {\it 3)} the
rotation period distribution for field stars of a given temperature
and age must be identical to that of equally aged open clusters
(metallicity differences, for instance, are ignored).  A rephrasing of
the first condition is that the star must be ``near the main
sequence'', because during post main sequence evolution, the
temperatures of stars change.  From $\approx$0.08 to $\approx$3\,Gyr,
the stars of interest in this work (3800--6200\,K,
0.5--1.2\,$M_\odot$) have temperatures constant to $\approx$1\%
\citep{Choi_2016}.  From $\approx$3--4\,Gyr, a 1.2\,$M_\odot$ star's
temperature drops from $\approx$6200\,K to $\approx$6000\,K
($\approx$3\%), because its outer layers expand as it begins hydrogen
shell burning as a subgiant.  We treat this issue in a somewhat
convoluted manner, discussed in ``bit 9'' below.

@@

{\it Absolute luminosity (bit 2)}---We calculated the absolute Gaia
DR3 $G$-band luminosity, ignoring reddening, using the reported
apparent $G$-band magnitude and parallaxes.  We flag stars with
$M_{\rm G} < 3.9$ or $M_{\rm G} > 8.5$, corresponding to main-sequence
spectral types earlier than $\approx$F8V or later than $\approx$M0.5V
\citep{Pecaut_2013}.

@@

{\it RUWE (bit 7)}---We inspected diagrams of the Gaia DR3 renormalized
unit weight error
(RUWE) as a function of other stellar parameters, %\footnote{\url{https://gea.esac.esa.int/archive/documentation/GDR3/Gaia_archive/chap_datamodel/sec_dm_main_source_catalogue/ssec_dm_gaia_source.html}})
and flag stars with RUWE$>$1.4 as possible binaries.  Such astrometric
outliers can be either bona fide astrometric binaries, or more often
are marginally resolved point sources for which the single-source PSF
model used in the Gaia pipeline induces spurious astrometric scatter.

@@

{\it Near the main-sequence (bit 9)}---Many stars observed by Kepler
are far from the main sequence: this can be seen in color--absolute
magnitude diagrams, and in diagrams of e.g.~surface gravity as a
function of temperature (Figure~\ref{fig:stellarprops}).  This is a
challenge for rotation-based age-dating because {\it i)} point sources
far from the main-sequence include unresolved photometric binaries,
and {\it ii}) effects other than magnetic braking influence the
rotation periods of bona fide single stars far from the main-sequence.
The added complications of reddening and metallicity make defining a
simple cut in any two dimensions challenging.  
We explored various options, and settled on the orange locus in the
$\log g$--$T_{\rm eff}$ plane shown in Figure~\ref{fig:stellarprops}.
While the details of how exactly this locus is constructed are
arbitrary and convoluted,\footnote{In practice, for K dwarfs we fitted an
$N^{\rm th}$-order polynomial to the set of stars for which $t_{\rm
B20,iso}/(+\sigma_{t,{\rm B20,iso}})\approx 3$, where
$t_{\rm B20,iso}$ was the isochronal age reported by
\citet{Berger_2020a_catalog}. We let the order of the fit vary from
$N$=1 to 10, minimized the Bayesian Information Criterion,
and found $N$=6.  For F and G dwarfs, this procedure failed to produce
a reasonable locus because of shifts in the relative precision of
the isochrone ages.  We therefore visually selected KOIs that appeared
to be near the ``main sequence'', fitted a separate polynomial to
their locations, and then stitched the resulting polynomials from the
F and G dwarfs to that of the K dwarfs.} the general 
aim is to exclude stars for which there is photometric evidence
that gyrochronology may not be reliable.  The rotation period
distribution of sources excluded in this manner includes a relatively
larger fraction of ``rapid rotators'' than the main sequence sources,
likely due to a combination of astrophysical and systematic effects.
See for instance Figure~8a of \citet{2022AJ....164..137K} or Figures~12 and 13
of \citet{2023ApJS..268....4F}.

@@

{\it How do we use the bit-flags?)}
An important decision for our subsequent analysis is how to combine
the above quality indicators in order to decide whether or not a star is suited for
gyrochronology.  By default, we assume that a star is suitable for
gyrochronology if none of bits zero through nine (inclusive) are
raised.

@@

{\it Candidate S/N (bit 1)}---the false positive rate increases
greatly toward the noise floor for planet detection in any transit survey \citep[e.g.][]{2002ApJ...564..495J}.  We
required a S/N in excess of Kepler's usual 7.1$\sigma$ floor, through a cut on the
``multiple event statistic'' (MES): we required ${\rm MES}$$>$10.

@@

In the planet sample, the cumulative KOI table included
\nnonfpcumkois\ confirmed and candidate planets, orbiting
\nnonfpcumkoihosts\ stars.  Assume that we apply all the same stellar
cuts, save for that on RUWE.  Then, require these confirmed and
candidate planets to have MES$>$10.  This combination of selection
criteria yield \nplwgyroagewithgrazingandhighruwe\ planets with
rotation-based ages orbiting
\nplhoststarwgyroagewithgrazingandhighruwe\ stars
(\nplhoststarwgyroagejusthighruwe\ of these stars with ``high'' RUWE).
If we additionally require the planets to be non-grazing ($b<0.8$)
based on the Markov Chain Monte Carlo (MCMC) fits from
\citet{Thompson_2018}, we are left with \nplwgyroagenograzing\ planets and
\nplhoststarwgyroagenograzing\ stars, respectively.
